---
title: 'gdv Report - Part 2'
subtitle: 'Visual Perception'
author: "Joseph Weibel"
date: "20.03.2021"
output: 
  pdf_document:
    keep_tex: true
    md_extensions: +raw_attribute
    
bibliography: bibliography.json
csl: apa.csl
geometry: left=1.5cm,right=1.5cm,top=1.9cm,bottom=1.9cm
header-includes:
  - \usepackage{graphicx}
  - \usepackage{wrapfig}
  - \usepackage{caption}
  - \captionsetup[figure]{font=tiny}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyfoot[L]{Joseph Weibel}
  - \fancyfoot[R]{gdv Report}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

original_plot <- knitr::knit_hooks$get("plot") 
knitr::knit_hooks$set(plot = function(x, options) {
  graphic <- original_plot(x, options)
  if(!is.null(options$wrapfigure)) {
    graphic <- str_replace(graphic, fixed("\\begin{figure}"), "")
    graphic <- str_replace(graphic, fixed("\\end{figure}"), "")
    begin <- sprintf('\\begin{wrapfigure}{%s}{%g\\textwidth}', options$wrapfigure[[1]], options$wrapfigure[[2]])
    return (c(begin, graphic, '\\end{wrapfigure}'))
  }
  return(graphic)
})


library(tidyverse)
library(zoo)
library(ggplot2)
library(RSwissMaps)
library(corrr)
source('./data.R')

raw_data <- load_data()
long_data <- raw_data %>%
  rename_with(~ gsub('^(.{2})\\.(.+)$', '\\1.kanton.\\2', .x), zh.berecht:ju.annahme) %>%
  pivot_longer(zh.kanton.berecht:ju.kanton.annahme, names_to = c('kanton', '.value'), names_pattern = "(.{2})\\.(.+)", values_drop_na = TRUE)

total_entries <- nrow(raw_data)
```

\vspace{10pt}

\begin{center}
Github Repository: https://github.com/josefweibel/ds-gdv-report
\end{center}

\newpage

# Visual Perception

Good visualisations allow the viewers to extract the key facts within a few seconds. The first decision to make for a good visualisation is to choose the right chart type. However, after that, there are many more decisions to take on how the data should be presented. Human brains can recognise objects and patterns subconsciously in a fraction of a second. Brains permanently collect information from the environment, filter it continuously, and the most relevant parts of it are processed further by conscious analysis [@http://zotero.org/users/7645537/items/GA6VT739]. This quality of extracting meaningful information in a short time is called pre-attentive processing, and by correctly designing a visualisation, data scientists can make use of it. 

## Gestalt Theory

Gestalt psychology helps to understand the abilities of human brains when it comes to visual perception. It emerged in the 20th century and averted from the then long-existing theory that humans recognise each component of a bigger picture individually. The Gestalt theory proposed that minds analyse patterns of components to understand the whole picture. The different types of patterns a human is able to inspect are the so-called Gestalt principles [@http://zotero.org/users/7645537/items/ZIUFQJ7Y].

One of these principles is the law of proximity. An individual automatically groups items being close together and separates them from items with more space between them even when there are lines and other elements to separate are missing [@http://zotero.org/users/7645537/items/RNUJTTDS]. Therefore, components belonging together should have less distance between them than to items from another group. Besides the proximity of components, the similarity can be used to group elements too. By giving certain items the same colour, viewers will instinctively assume they belong together. Besides colours, this also works when using the same shape, shading or other properties [@http://zotero.org/users/7645537/items/ZIUFQJ7Y] and will often be used in data visualisation to express similarity between specific markers. 

The so-called law of closure says that visualised objects with gaps do not stop individuals from seeing the complete objects. A brain will perceive many standalone dots as the line they are forming [@http://zotero.org/users/7645537/items/RNUJTTDS]. For instance, in data visualisation, scatter plots aim to show a correlation between two variables by showing many points that form a line when the variables correlate. Also, humans automatically try to identify symmetrical objects when viewing a picture. Instead of seeing complex figures, they use symmetry to extract known simple forms. The law of similarity helps to identify parts in a graphic that will be inspected for symmetrical objects [@http://zotero.org/users/7645537/items/RNUJTTDS]. The following principle, the law of common fate, is relevant for interactive visualisations. It states that humans distinguish between the objects moving in the same direction and the ones which are not moving or move in another direction. The items not behaving like the others will be perceived actively by the human's mind [@http://zotero.org/users/7645537/items/6GJ2AUHC].

Humans require the law of continuity to understand line charts with multiple cutting lines. Their eyes will follow a line on the point of intersection on the smoothest path. Sudden bends are not expected, and such can confuse the viewer [@http://zotero.org/users/7645537/items/RNUJTTDS]. Brains do not only analyse visualised elements. They also interpret the whitespace around elements. This law is often deliberately used for visual illusions but can also be used by product designers and data scientists to highlight specific items [@http://zotero.org/users/7645537/items/RNUJTTDS].

## Bertin's Characteristics

These theoretical principles are all relevant when designing data visualisations. French cartographer Jacques Bertin determined seven concrete visual variables for markers based on these Gestalt principles to create meaningful maps and other graphics [@http://zotero.org/users/7645537/items/286WNG6C]. Bertin categorised these characteristics into four groups. All variables except shape and orientation are selective, meaning the viewer can determine the different groups at first glance, whereby associative variables like shape and colour hue can determine the group of a single marker. If the viewer can determine an order of the visual variable's values, they get classified as ordinal. Ordinal values are colour saturation and size but not shapes. Finally, if it is possible to extract a value from the variable, it is categorised as quantitative, like the size but not the colour hue [@http://zotero.org/users/7645537/items/ZSPWKPQD]. 

### Location

```{r location_scale, fig.asp=0.2, out.width="50%", fig.show = "hold", fig.cap=location_scale_caption}
location_scale_caption <- 'Line charts showing voter participation over time with different y scales.'
plot_linechart <- function (y_start = 0, y_end = 100) {
  rolling_period = 5
  data <- as_tibble(raw_data %>%
    filter(!is.na(bet)) %>%
    group_by(jahr) %>%
    summarise(avg_bet = mean(bet), .groups = 'drop') %>%
    ungroup() %>%
    rollapply(rolling_period, mean, align = 'right'))
  ggplot(data, mapping = aes(x = jahr, y = avg_bet)) +
    geom_line() +
    xlim(1879, 2020) +
    ylim(y_start, y_end) +
    labs(
      title = 'Voter Participation over Time',
      caption = sprintf('Rolling period over last %d years', rolling_period),
      y = 'Participation in Percent',
      x = 'Voting Date'
    )
}

plot_linechart(35, 75)
plot_linechart()
```

```{r location_periodic1, fig.asp=0.3, out.width=".1\\textwidth", wrapfigure=list("R", 0.2), fig.cap=location_periodic1_caption}
location_periodic1_caption <- 'Line chart using a cartesian coordinate system.'
plot_periodic_linechart <- function (use_polar = 0) {
  data <- raw_data %>%
    group_by(monat) %>%
    summarise(count = n(), .groups = 'drop') %>%
    ungroup()
  
  if (use_polar) {
    data <- data %>%
      add_row(monat = 0, count = data[which(data$monat == 12), ]$count)
  }

  plot <- ggplot(data, mapping = aes(x = monat, y = count)) +
    geom_line() +
    scale_x_continuous(breaks = 1:12) +
    labs(
      title = 'Number of Referendums per Month',
      y = 'Number of Referendums',
      x = 'Month'
    )
  
  if (use_polar) {
    plot <- plot + coord_polar()
  }
  
  return(plot)
}

plot_periodic_linechart()
```

Any visualisation requires the positioning of markers, and in most cases, values will be placed on a conventional 2D Cartesian coordinate system [@http://zotero.org/users/7645537/items/Y2NS2S2E]. Certain expectations exist on which type of values are placed on which axis, like time values are usually placed on the y axis. The scales should always start at the zero-point to avoid misinterpretation when numeric values are placed on axes. The voter participation trend in Figure 1 on the left side implies more significant deviations over time than they actually were when comparing with the right plot having a y-axis from 0 to 100.

```{r location_periodic2, fig.asp=0.3, out.width=".2\\textwidth", wrapfigure=list("R", 0.2), fig.cap=location_periodic2_caption}
location_periodic2_caption <- 'Line chart using a polar coordinate system.'
plot_periodic_linechart(1)
```

When both scales are of the same unit, it is good advice to have equal segmentations on both axes. However, the scales do not always have to be linear. Logarithmic scales can be used to visualise an exponential slope so that the focus lies on changes in the increase of the value. However, such scales can never show the value zero. There are also alternatives to the cartesian coordinate system. The polar coordinate system can show periodic data like months of a year [@http://zotero.org/users/7645537/items/Y2NS2S2E]. Figure 2 and 3 show the same data. The number of referendums per month is shown on a regular cartesian and a polar coordinate system. On both, it becomes evident, referendums are usually held during the same time of the year, and differences between some months are enormous. 

```{r shape, fig.asp=0.2, out.width=".2\\textwidth", wrapfigure=list("R", 0.2), fig.cap=shape_caption}
shape_caption <- 'Chart using different shapes.'

bet_trend_values <- c('increasing', 'constant', 'decreasing')
canton_voter_participation_data <- long_data %>%
  filter(jahr == 2019) %>%
  group_by(kanton) %>%
  summarise(
    avg_bet_2019 = mean(kanton.bet), 
    avg_berecht_2019 = mean(kanton.berecht), 
    .groups = 'drop'
  ) %>%
  ungroup() %>%
  mutate(
    avg_bet_2018 = long_data %>%
      filter(jahr == 2018) %>%
      group_by(kanton) %>%
      summarise(avg_bet_2018 = mean(kanton.bet), .groups = 'drop') %>%
      ungroup() %>%
      select(avg_bet_2018),
    bet_trend = case_when(
      avg_bet_2019 - avg_bet_2018 > 0 ~ 'increasing',
      avg_bet_2019 - avg_bet_2018 < 0 ~ 'decreasing',
      TRUE ~ 'constant'
    ),
    bet_trend_compulsory = ifelse(kanton %in% c('sh'), paste0(bet_trend, ' (compulsory)'), bet_trend)
  ) %>%
  mutate(bet_trend = factor(bet_trend, levels = bet_trend_values, labels = bet_trend_values))

plot_shape <- function () {
  ggplot(canton_voter_participation_data, mapping = aes(x = toupper(kanton), y = avg_bet_2019, shape = bet_trend)) +
    geom_point() +
    scale_shape_manual(values = c('decreasing' = 6, 'constant' = 5, 'increasing' = 2)) +
    ylim(0, 100) +
    labs(
      title = 'Average Voter Participation 2019 per Canton',
      y = 'Voter Participation',
      x = 'Canton',
      shape = 'Compared to 2018'
    )
}

plot_shape()
```

### Shape and Size

Every marker is represented by a shape. A shape is usually a line, a point or a bar depending on the chart. However, there are more variations for each chart type to categorise values. As an alternative to points, there are arrows, squares, triangles and circles, which are then mapped to a category on a legend [@http://zotero.org/users/7645537/items/FBBRGNJV].

```{r size, fig.asp=0.2, out.width=".2\\textwidth", wrapfigure=list("R", 0.2), fig.cap=size_caption}
size_caption <- 'Bar chart using different sizes.'
plot_size <- function (fill = NULL, highlight = FALSE) {
  fill_colours = c('decreasing' = '#d95f02', 'constant' = '#7570b3', 'increasing' = '#1b9e77')
  
  if (highlight) {
    fill_colours = c(
      'decreasing (compulsory)' = '#de2d26', 
      'constant (compulsory)' = '#3182bd', 
      'increasing (compulsory)' = '#31a354',
      'decreasing' = '#fc9272', 
      'constant' = '#9ecae1', 
      'increasing' = '#a1d99b'
    )
  }
  
  max_berecht <- max(canton_voter_participation_data$avg_berecht_2019)
  
  ggplot(canton_voter_participation_data) +
    geom_bar(mapping = aes(x = toupper(kanton), y = avg_bet_2019, fill = fill), width = canton_voter_participation_data$avg_berecht_2019 / max_berecht * 1.7, stat = 'identity') +
    ylim(0, 100) +
    scale_fill_manual(values = fill_colours) +
    labs(
      title = 'Average Voter Participation 2019 per Canton',
      y = 'Voter Participation',
      x = 'Canton',
      size = 'Number of Eligible Voters',
      fill = 'Compared to 2018'
    )
}

plot_size()
```

The size of the item to show should not be neglected. Obviously, the bars' lengths in a bar chart are essential, but their width can be used to visualise an additional variable if needed and suitable. The example on the right shows the same data as in the last plot. However, each column's width represents the number of eligible voters in each canton to indicate the viewer the absolute voter participation as the differences are significant.


```{r colour1, fig.asp=0.2, out.width=".2\\textwidth", wrapfigure=list("R", 0.2), fig.cap=colour1_caption}
colour1_caption <- 'Bar chart using different colour hues.'
plot_size(fill = canton_voter_participation_data$bet_trend)
```

### Colour

Colour is an essential characteristic of data visualisations. However, they can have different purposes. The shading (colour hue) is selective and associative as the viewer can distinguish between the various colour hues and can categorise them using a legend. Thus, hues are used to group values using a categorical data variable. Colour palettes for that purpose contain a finite number of distinct but matching colours, whereby none of them stands out. That is important to prevent the viewer from seeing order in them, which should not be the case since all groups should be treated equally [@http://zotero.org/users/7645537/items/Y2NS2S2E]. Consider the plot on the right side for an example. When the participation increased in 2019 compared to 2018, the bars are painted greenish; otherwise reddish. 

```{r colour2, fig.asp=0.2, out.width=".2\\textwidth", wrapfigure=list("R", 0.2), fig.cap=colour2_caption}
colour2_caption <- 'Bar chart using different colour hues and values.'
plot_size(fill = canton_voter_participation_data$bet_trend_compulsory, highlight = TRUE) +
  labs(title = 'Canton Schaffhausen the only Canton with Voting Obligation')
```

Often it is desired to let the viewer use the colours to compare different continuous values. In these cases, the colours must indicate a natural order which can be resolved using different brightnesses as an example. They can be all of the same hues or use multiple following typical gradients (e.g. dark red to light yellow). There are also dedicated palettes to present deviations in one or two directions. These palettes contain light colours for slight deviation and dark colours for high deviations. The colours must also indicate a natural order. Finally, colours can be used to bring the focus of the viewer to specific data points. The highlighted data points are painted in darker or more saturated colours than the others, whereby colour hues can match both colour schemes [@http://zotero.org/users/7645537/items/Y2NS2S2E]. The example on the right highlights all cantons where voting is compulsory (only Schaffhausen) and uses colour hues to show the participation trend. When choosing colours, it is vital to consider particular which are distinguishable by colourblind people. Predefined colour palettes are available at colorbrewer2.org, where all mentioned aspects were considered. It is also possible to use different texture on the markers or change the transparency, which is advisable when markers can overlap.

### Orientation

The orientation of a shape can indicate another value. For instance, an arrow can change its meaning depending on how it is rotated (see Figure 4). However, when it comes to 2D maps, a study proves that the orientation has the most negligible benefit for the viewer of all these visual variables [@http://zotero.org/users/7645537/items/ZSPWKPQD].

\newpage

# Appendix

## Figures

### Figure 1: `r location_scale_caption`

```{r appendix-figure-location-scale, out.width="50%", fig.show = "hold", fig.width=6}
plot_linechart(35, 75)
plot_linechart()
```

### Figure 2: `r location_periodic1_caption`

```{r appendix-figure-location-periodic-1, fig.width=12}
plot_periodic_linechart()
```

### Figure 3: `r location_periodic2_caption`

```{r appendix-figure-location-periodic-2, fig.width=12}
plot_periodic_linechart(1)
```

### Figure 4: `r shape_caption`

```{r appendix-figure-shape, fig.width=12}
plot_shape()
```

### Figure 5: `r size_caption`

```{r appendix-figure-size, fig.width=12}
plot_size()
```

### Figure 6: `r colour1_caption`

```{r appendix-figure-colour1, fig.width=12}
plot_size(fill = canton_voter_participation_data$bet_trend)
```

### Figure 7: `r colour2_caption`

```{r appendix-figure-colour2, fig.width=12}
plot_size(fill = canton_voter_participation_data$bet_trend_compulsory, highlight = TRUE) +
  labs(title = 'Canton Schaffhausen the only Canton with Voting Obligation')
```


\newpage

# References
